FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-02-03 manuscript.Rmd
do something with felix's comments

As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors such as adherence to pandemic mitigation methods (e.g., ), pandemic-related social disruption (e.g., panic buying), or pandemic-related distress and related problems (e.g., anxiety) [@taylor2022psychology].
The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Making theories Findable would allow researchers to easily identify relevant theories
and ground their hypotheses in established theoretical foundations.
It further increases the impact and reuse potential of theories across disciplines,
either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
or through analogical modeling.
In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of intelligence [@vandermaasDynamicalModelGeneral2006], and the Eysenck model of atomic magnetism has inspired a network theory of depression [@cramerMajorDepressionComplex2016].
Findability also enables meta-research on theories, 
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
In a similar way, it would become much easier to explicitly compare different theories of a specific phenomenon,
or to study structural properties of theories.

The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
First, this requires assigning a globally unique and persistent identifier, or DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artefacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena [REF Noah Denny].
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

@@

At present, there are several impediments to theories' accessibility.
To the extent that theories are still contained within papers,
paywalls erected by commercial publishers constitute a barrier.
Open Access publishing thus increases the accessibility of all academic output, including theory.
A second impediment is more indirect:
While open access publishing increases practical access to theories,
accessibility also requires clear and explicit communication.
<!-- NvD: I am not sure if we need to go into this much detail as presented in the rest of this Section. My suggestion is to remove most of it. I'll indicate which elements I suggest keeping -->
This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. <!-- NvD: I suggest keeping this -->
At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].<!-- NvD: I suggest keeping this -->
It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
which holds that every communicative utterance has multiple alternative translations, with no *objective* means of choosing the correct one.
It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
This places a theoretical upper bound on theories' ability to be understood.

Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
The Kuhnian notion of "normal science", conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding [@kuhnStructureScientificRevolutions2009].
From a pragmatic perspective, these considerations indicate that,
when striving to make theory accessible, <!-- NvD: I suggest keeping this -->
it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions), <!-- NvD: I suggest keeping this -->
while acknowledging that accessibility exists on a spectrum, <!-- NvD: I suggest keeping this -->
and that it is impossible to eliminate all ambiguity.<!-- NvD: I suggest keeping this -->
Rather, it may benefit scientific discourse to anticipate misunderstanding,<!-- NvD: I suggest keeping this -->
and use it to drive further explication of theory.<!-- NvD: I suggest keeping this -->
In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)<!-- NvD: I suggest keeping this -->
will advance its Accessibility.<!-- NvD: I suggest keeping this -->

@@

Aside from legal conditions for reuse, there are also social considerations.
For example, while a CC0 license does not mandate attribution,
it is nonetheless essential that scholars comprehensively cite theory and related works to comply with established norms of attribution and comprehensive citation [@aalbersbergMakingScienceTransparent2018].
Another way to guide the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
It is also possible to create or adopt a "Code of Conduct" which prescribes

@@

Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the *empirical cycle*,
we decided to use it as an example for this tutorial.
The resulting FAIR theory is available at <https://doi.org/10.5281/zenodo.14552329>.
The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969.
Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking".
We can thus consider it a meta-theory of theory construction.
<!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
The "empirical cycle" theory presented by De Groot consists of a series of natural language statements:

@@

This implementation describes the model as a directed graph.
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].

We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

### What's in a README?

The readme should contain information to help people get started with using your FAIR theory.
We suggest the following elements:

* Title, prefaced with `# FAIR theory: The Theory's Name`
* Description: A plain-text description of the theory and its scope
* Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use". From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)". We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
* Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
* License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*.
* Citing this work: Tell users how to cite the theory. Note that this section is redundant with the Zenodo archive, which has a preferred citation field. The disadvantage of redundant information is that you may have to maintain this section of the README going forward. The advantage is that documenting related works in the README makes it more readily accessible to users. We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
* Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Again, this is redundant with metadata entered in Zenodo (step 5). We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here:

```
This repository contains an implementation of the "empirical cycle",
a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.

> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
Foundations of inference and research in the behavioral sciences.
De Gruyter Mouton. https://doi.org/10.1515/9783112313121
```

>>>>

As an applied example, consider a comprehensive theory of disease spread and pandemics which covers various psychological factors
such as adherence to infection prevention protocols (e.g., social distancing),
pandemic-related behavior (e.g., panic buying),
and pandemic-related distress
[@taylor2022psychology].
The theory may encompass a particular transmission _model_ for disease spread including precise parameters for the process of infection (e.g., social distance, average duration of encounters, ventilation) and incubation times.

@@

Making theories Findable would allow researchers to easily identify relevant theories
and ground their hypotheses in established theoretical foundations.
It further increases the impact and reuse potential of theories across disciplines,
either through direct application (where one discipline stumbles upon a problem that is already well-understood in another discipline),
or through analogical modeling.
In analogical modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of intelligence [@vandermaasDynamicalModelGeneral2006], and the Eysenck model of atomic magnetism has inspired a network theory of depression [@cramerMajorDepressionComplex2016].
Findability also enables meta-research on theories, 
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
In a similar way, it would become much easier to explicitly compare different theories of a specific phenomenon,
or to study structural properties of theories.

The four Findability criteria are applicable to theory with only minor adjustments, see Table \@ref(tab:tabfair).
First, this requires assigning a globally unique and persistent identifier, such as a DOI, to each theory (F1).
Of the many services that provide DOIs for scientific information artefacts,
Zenodo and the Open Science Framework are commonly used in psychology.
Second, Findable theory is described with rich metadata	(F2).
This includes citation metadata (e.g., referencing a scientific paper that documents the theory, or a psychometric paper that operationalizes specific constructs).
It might further include domain-specific metadata, such as a reference to a taxonomy of psychological constructs [@boscoMetaBUSVehicleFacilitating2017],
ontology [@guyonMeasurementOntologyEpistemology2018],
or catalog of psychological phenomena [REF Noah Denny].
Metadata should also include identifiers for all the versions of the theory it describes	(F3);
Zenodo handles this by default by providing an overarching DOI for an information artifact which subsumes the DOIs of that artifact's versions.

@@

At present, there are several impediments to theories' accessibility.
To the extent that theories are still contained within papers,
paywalls erected by commercial publishers constitute a barrier.
Open Access publishing thus increases the accessibility of all academic output, including theory.
A second impediment is more indirect:
While open access publishing increases practical access to theories,
accessibility also requires clear and explicit communication.
<!-- NvD: I am not sure if we need to go into this much detail as presented in the rest of this Section. My suggestion is to remove most of it. I'll indicate which elements I suggest keeping -->
This property of good theories has been dubbed "discursive survival [...], the ability to be understood" [@guestWhatMakesGood2024]. 
At present, psychological theories are often ambiguous, rendering them difficult to understand [@frankenhuisStrategicAmbiguitySocial2023].
It is important to acknowledge the *indeterminacy of translation* [@quineReasonsIndeterminacyTranslation1970]:
which holds that every communicative utterance has multiple alternative translations, with no *objective* means of choosing the correct one.
It follows that an idea cannot be formalized to the point that it becomes unambiguously interpretable.
This places a theoretical upper bound on theories' ability to be understood.

Successful communication requires shared background knowledge between sender and receiver [@vogtFAIR20Extending2024].
The Kuhnian notion of "normal science", conducted within the context of a shared paradigm, provides shared background knowledge to facilitate mutual understanding [@kuhnStructureScientificRevolutions2009].
From a pragmatic perspective, these considerations indicate that,
when striving to make theory accessible, 
it is important to be as explicit as possible (e.g., about assumptions  and ontological definitions), 
while acknowledging that accessibility exists on a spectrum, 
and that it is impossible to eliminate all ambiguity.
Rather, it may benefit scientific discourse to anticipate misunderstanding,
and use it to drive further explication of theory.
In sum, efforts to communicate theory clearly, with as few dependencies on shared background knowledge as possible, including by formalization, explication of assumptions,
and cross-references to resources that provide relevant context (papers, ontologies, macro-theories, theories of measurement)
will advance its Accessibility [@langeChecklist2025].

@@

Aside from legal conditions for reuse, there are also social considerations.
For example, while a CC0 license does not legally mandate attribution,
the norms of good scientific practice require that scholars comprehensively cite theory and related works [@aalbersbergMakingScienceTransparent2018].
Another way to guide the social process of (diffuse) collaboration is to include a "README" file in the theory repository, which informs users about the ways in which they can reuse and contribute to a FAIR theory.
It is also possible to create or adopt a "Code of Conduct" which prescribes behavioral norms for contributors and users of a theory.

@@

We will use the *empirical cycle* as a running example for this tutorial.
The empirical cycle, described on page 28 of @degrootMethodologyFoundationsInference1969,
is a kind of meta-theory of theory construction.
The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
<!-- Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking". -->
<!-- We can thus consider it a meta-theory of theory construction. -->
<!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
The original theory consists of a series of natural language statements:

@@

This implementation describes the model as a directed graph.
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html)
We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

<!-- ### What's in a README? -->

<!-- The readme should contain information to help people get started with using your FAIR theory. -->
<!-- We suggest the following elements: -->

<!-- * Title, prefaced with `# FAIR theory: The Theory's Name` -->
<!-- * Description: A plain-text description of the theory and its scope -->
<!-- * Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use". From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)". We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph. -->
<!-- * Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*. -->
<!-- * License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*. -->
<!-- * Citing this work: Tell users how to cite the theory. Note that this section is redundant with the Zenodo archive, which has a preferred citation field. The disadvantage of redundant information is that you may have to maintain this section of the README going forward. The advantage is that documenting related works in the README makes it more readily accessible to users. We suggest a compromise: to retain this section, but refer the reader to the Zenodo page. -->
<!-- * Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Again, this is redundant with metadata entered in Zenodo (step 5). We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here: -->

<!-- ``` -->
<!-- This repository contains an implementation of the "empirical cycle", -->
<!-- a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works. -->

<!-- > De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology: -->
<!-- Foundations of inference and research in the behavioral sciences. -->
<!-- De Gruyter Mouton. https://doi.org/10.1515/9783112313121 -->
<!-- ``` -->
